Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-10-10 paper/paper.tex
remove github footnote link
Haskell for inductive/coindutive type

\subsubsection{Functors in Haskell}
We can declare the class of types which are functors. This is done by imposing that a function \mintinline{haskell}{f : * -> *} comes with a functorial action given by function \mintinline{haskell}{fmap}:
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

Now the type of lists can be defined as follows along with their instance of the class \mintinline{haskell}{Functor}:
\begin{minted}{haskell}
data List a = Nil | Cons a (List a)

@@

\subsubsection{Inductive Types and Catamorphisms in Haskell }
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
data ListF a x = Nil | Cons a x deriving Functor
\end{minted}

\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}
\begin{minted}{haskell}
foldr :: (a -> b -> b) -> b -> [ a ] -> b
foldr f e [ ]    = e
foldr f e (x:xs) = f x (foldr f e xs)
\end{minted}

@@

\subsubsection{Coinductive Types and Anamorphisms in Haskell }

\begin{minted}{haskell}
newtype CoFix f = OutOp {out :: f (Fix f )}
\end{minted}

\begin{minted}{haskell}
ana :: Functor b => (a -> b a) -> a -> CoFix b
ana coalg = OutOp . fmap (ana coalg) . coalg
\end{minted}

@@

\subsubsection{Haskell Implementation of Hylomorphisms}

which is very useful in list processing. As a fold, The frequently used function map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}
 and can be

>>>>

\subsubsection{Functors in Haskell}
It is often helpful for functional programmers to think of categorical
constructions in terms of Haskell thereby assuming that Haskell types are sets
and programs are functions~\cite{DanielssonHJG06}.

We can declare the class of types which are functors. This is done by imposing
that a function \mintinline{haskell}{f : * -> *} comes with a functorial action
given by function \mintinline{haskell}{fmap}:
\begin{minted}{haskell}
class Functor f where
  fmap :: (a -> b) -> f a -> f b
\end{minted}

We now define the type of lists as usual and then show that it is an instance of the class \mintinline{haskell}{Functor} by implementing the \mintinline{haskell}{fmap} program:
\begin{minted}{haskell}
data List a = Nil | Cons a (List a)

@@

\subsubsection{Inductive Types and Catamorphisms in Haskell }
We now implment the least fixed-point of a functor \mintinline{haskell}{f}.
Inuitively this is the type of \mintinline{haskell}{f}-branching trees of finite depth:
\begin{minted}{haskell}
newtype Fix f = In {inOp :: f (Fix f )}
\end{minted}
Note that since Haskell is a lazy language with general recursion this type can
also represent infinite trees. However, the point of using recursion schemes is
precisely that of avoid using general recursion in the first place.

A catamorphism can be readily implemented as follows
\begin{minted}{haskell}
cata :: Functor f => (f a -> a) -> Fix f -> a
cata alg = alg . fmap (cata alg) . In
\end{minted}

As an example of how to construct inductive types we define the functor $X \mapsto 1 + A \times X$ as data type parameterized by a type:
\begin{minted}{haskell}
data ListF a x = Nil | Cons a deriving Functor
newtype List a = (Fix ListF) a
\end{minted}
It should be easy now to see that a foldr is precisely a catamorphism over the
type \mintinline{haskell}{List} which we have just defined.  For example,
summing up the elements of a list can be implemented using a catamorphism as
follows
\begin{minted}{haskell}
sum :: Fix (ListF Int) -> Int
sum xs = cata (\case { Nil -> 0; Cons n m -> n + m } )  xs
\end{minted}
In the program above, the code inside the lambda case is precisely the \mintinline{haskell}{ListF Int}-algebra for the catamorphism.

@@

\subsubsection{Coinductive Types and Anamorphisms in Haskell }
Similarly to the inductive case, we can implement coinductive types and
catamorphisms as follows:
\begin{minted}{haskell}
newtype CoFix f = OutOp {out :: f (Fix f )}
\end{minted}
As mentioned above, this type is similar to the data type
\mintinline{haskell}{Fix} we already defined. However, this time we are going to
use this assuming that it is the coinductive fixed-point by defining an
anamorphism on it:
\begin{minted}{haskell}
ana :: Functor b => (a -> b a) -> a -> CoFix b
ana coalg = OutOp . fmap (ana coalg) . coalg
\end{minted}

For example, the streams of natural numbers can now be defined using an anamorphism
\begin{minted}{haskell}
nat :: CoFix (ListF Int)
nat = ana (\x -> Cons x (x+1)) 0
\end{minted}
where the argument to the anamorphism is the
\mintinline{haskell}{ListF Int}-coalgebra.

@@

\subsubsection{Haskell Implementation of Hylomorphisms}
which is very useful in list processing. As a fold, The frequently used function
map is also a fold:
\begin{minted}{haskell}
map :: (a -> b) -> [a] -> [b]
map f = foldr (\x xs -> (f x):xs) []
\end{minted}
 and can be

\begin{minted}{haskell}
hylo :: (c -> f c) -> (a -> f a) -> c -> a
hylo c a = a . fmap (hylo c a). c
\end{minted}
